---
title: "RFLOMICS data management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RFLOMICS-data-management}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>"
)

```

RFLOMICS is a R package that provides a complete workflow for the
analysis of multi-omics data within complex experimental designs
framework.

To ensure an efficient data management (statistical design, data,
analysis parameters and results) we used the MultiAssayExperiment and
SummarizedExperiment classes that we extended to fix a given
organization for the slot metadata.

# The RflomicsMAE and RflomicsSE classes:

The RflomicsMAE (which extend the MultiAssayExperiment class) will store
the statistical design and multi-omics data integration analysis
parameters and results in the metadata slot. It will contain an
ExperimentList of RflomicsSE (which extend SummarizedExperiment class)
storing single-omics data and analysis parameters and results in the
metadata slot.

In this vignette, we will use an example to show how the metadata slots
of the two RflomicsMAE and RflomicsSE objects are organised and can be
accessed with the progress of the analysis workflow.

RFLOMICS workflow is structured into three main parts:

-   The first one involves loading the experimental design and omics
    data, setting up the statistical model, and defining contrasts
    (hypotheses) associated with the biological issues. The design List
    of the metadata slot of the RflomicsMAE object stores this
    statistical settings (in green in Fig 1).

![Fig 1: Organization of the metadata slot of the RflomicsMAE
object](Images/DM_RflomicsMAE-design.png){width="80%"}

-   The second part allows for complete single-omics analysis for each
    dataset. Both raw and processed omic datasets are stored in the
    ExperimentList as RflomicsSE. For each analysis step, settings and
    results are stored (in green in Fig 2)

    -   i- Quality control to check for batch effects or identify
        outlier samples:

    -   ii- The DataProcessing List of the metadata slot keeps
        pre-processing which may include filtering, normalization,
        and/or transformation.

    -   iii- The DiffExpAnal List of the metadata slot stores the
        Differential expression analysis settings and results.

    -   vi- The CoExpAnal List of the metadata slot stores co-expression
        analysis on differentially expressed entities settings and
        results.

    -   vii- DiffExpEnrichAnal and CoExpEnrichAnal List of the metadata
        slot keeps Functional enrichment analysis results and settings
        for respectively the Differential expression analysis and
        CoExpression analysis.

![Fig 2: Organization of the metadata slot of the RflomicsSE
objects](Images/DM_RflomicsSE-overview.png){width="100%"}

-   The third part is the multi-omics integration, with two proposed
    methods: a supervised approach using the MixOmics package (DIABLO),
    and an unsupervised approach using MOFA2. (settings and results
    stored in RflomicsMAE object)

![Fig 3: storage of multi-omics integration settings and results in the
metadata slot of the RflomicsMAE
object](Images/DM_RflomicsMAE-integration.png){width="80%"}

As you will see, this analysis workflow is implemented as methods of the
two classes (RflomicsMAE and RflomicsSE), providing a generic interface
regardless of the type of omics data being analyzed (RNAseq, proteomics,
metabolomics, etc.). This methods will return the same class of object
(RflomicsSE or RflomicsMAE) as the entry object type with the metadata
slot enriched with the parameters and results. We also defined getter
and setter to access this objects.

# Setter and getter for the RflomicsMAE and RflomicsSE metadata slots:

## Instantiate a RflomicsMAE from MAE:

In this example we will Instantiate a RflomicsMAE object from an existing MAE object and a design data.frame.

```{r, echo=FALSE}
library(RFLOMICS)
```

```{r ecoseed2mae, echo=FALSE}

data(ecoseed)

## Create the SummarizedExperiment objects from the 3 data.frames 

RNAtest.SE <- SummarizedExperiment( assays  = as.matrix(ecoseed$RNAtest)) 

protetest.SE <- SummarizedExperiment( assays  = as.matrix(ecoseed$protetest)) 

metatest.SE <- SummarizedExperiment( assays  = as.matrix(ecoseed$metatest)) 

## Combine to a named list and call the ExperimentList constructor function 

assayList <- list(RNAtest=RNAtest.SE, protetest=protetest.SE, metatest=metatest.SE)

ExpList <- ExperimentList(assayList)

## Create the sampleMap

RNAtest.map <- data.frame(
    primary = colnames(ecoseed$RNAtest),
    colname = colnames(ecoseed$RNAtest),
    stringsAsFactors = FALSE)

protetest.map <- data.frame(
    primary = colnames(ecoseed$protetest),
    colname = colnames(ecoseed$protetest),
    stringsAsFactors = FALSE)

metatest.map <- data.frame(
    primary = colnames(ecoseed$metatest),
    colname = colnames(ecoseed$metatest),
    stringsAsFactors = FALSE)

maplist <- list(RNAtest=RNAtest.map, protetest=protetest.map, metatest=metatest.map)

sampMap <- listToMap(maplist)

## Create a colData

colDat <- ecoseed$design

## Create an example phenotype data
mae <- MultiAssayExperiment(experiments = ExpList, colData = colDat, sampleMap = sampMap, metadata = list("DataRef"="https://www.uibk.ac.at/botany/ecoseed/home/"))

# save(mae, file="mae.RData")
```

```{r createRlomisMAE, echo = TRUE}
# à venir
# testObject <- createRlomisMAE(
#  projectName = "Tests",
#  omicsData=ecoseed$mae, 
#  omicsNames  = c("RNAtest", "metatest", "protetest"),
#  omicsTypes  = c("RNAseq","metabolomics","proteomics"),
#  ExpDesign   = colData(ecoseed$mae),
#  factorRef   = ecoseed$factorRef)

# now
testObject <- createRflomicsMAE(
    projectName = "Tests",
    omicsData   = list(ecoseed$RNAtest, ecoseed$metatest, ecoseed$protetest),
    omicsNames  = c("RNAtest", "metatest", "protetest"),
    omicsTypes  = c("RNAseq","metabolomics","proteomics"),
    ExpDesign   = ecoseed$design,
    factorRef   = ecoseed$factorRef)

# Note the ".raw" suffix that was added to each of the names. This serves as a security measure for the interface pipeline. For this example, we won't use it.

names(testObject) <- gsub("[.]raw", "", names(testObject))
```

### Statistical settings

- set the model formulae

```{r designMetadata, echo = TRUE}
# generate the model formulae
form <- generateModelFormulae(testObject)

# set the model formulae
testObject <- setModelFormula(testObject, form[2])

# get the m model formulae
getModelFormula(testObject)
```

- set contrasts:

```{r contr, echo =TRUE}
# generate the possible contrasts
possibleContrasts <- generateExpressionContrast(testObject)

# set the contrasts
testObject <- setSelectedContrasts(testObject, 
                                   contrastList = possibleContrasts$averaged[1:3])

# get the contrasts
getSelectedContrasts(testObject)
```


## Set and get Single-Omic analysis settings and results

- Pre-processing of data: normalization and transformation settings and results

```{r runNorm, echo=TRUE}

# Noramlize and transform proteomics and metabolomics data 
testObject <- runNormalization(testObject, 
                               SE.name = "metatest", 
                               normMethod = "median") |>
    runTransformData(SE.name = "metatest", transformMethod = "log2") |>
    runTransformData(SE.name = "protetest", transformMethod = "log2") |>
    runOmicsPCA(SE.name = "metatest")

# Access the normalization settings
getNormSettings(testObject[["metatest"]])

# Noramlize RNAseq data
testObject <- filterLowAbundance(testObject, SE.name = "RNAtest", 
                                 filterMethod = "CPM",
                                 filterStrategy = "NbReplicates",
                                 cpmCutoff = 5) |>
    runNormalization(SE.name = "RNAtest", normMethod = "TMM")

getFilteredFeatures(testObject[["RNAtest"]])[1:10]
```

It is important to note that none of the data are directly transformed
in the assay, except if specifically asked by the user with setting the
argument `modifyAssay` to TRUE. This way, you can always change your
mind and test several normalization or transformation processes.

You can access all results and plot them.

```{r plotPCA, echo=TRUE}
plotOmicsPCA(testObject[["metatest"]], raw = "raw")
plotOmicsPCA(testObject[["metatest"]], raw = "norm")
```

```{r plotDataDistribution, echo=TRUE}
plotDataDistribution(testObject[["metatest"]], raw = TRUE)
plotDataDistribution(testObject[["metatest"]], raw = FALSE)
```

- Differential analysis: settings and results

```{r diff, echo=TRUE}
# run differential analysis on RNAseq data
testObject <- runDiffAnalysis(testObject, SE.name = "RNAtest")

# access to the settings
getDiffSettings(testObject, SE.name = "RNAtest")

# plot results as volcano
plotDiffAnalysis(testObject, 
                 SE.name = "RNAtest", 
                 contrastName = "(temperatureMedium - temperatureLow) in mean", 
                 typeofplots = "volcano")

# as boxplot for a given gene
plotBoxplotDE(testObject[["RNAtest"]], 
              features = "AT4G04810")

```

## Set and get Multi-Omics integration analysis settings and results 

The multi-omics step in RFLOMICS takes two steps: the preparation of the
object, based on the user-selected method, and the actual run of the
selected method. This is done using `prepareForIntegration` and
`runOmicsIntegration`. In this example, we will use only the proteomics
and metabolomics dataset with the mixOmics package (block.plsda
function).

`prepareForIntegration` does not return a RflomicsMAE object, it returns
the corresponding type to be an input for either MOFA2 or mixOmics
functions.

```{r, echo=TRUE}
listIntegration <- prepareForIntegration(testObject, 
                                         omicsNames = c("metatest", "protetest"),
                                         method = "mixOmics"
)

testObject <- runOmicsIntegration(object = testObject, 
                                  preparedObject = listIntegration, 
                                  method = "mixOmics", 
                                  selectedResponse = "temperature", 
                                  ncomp = 3)
```

Visualizing results is made through the original package functions.

```{r , echo=TRUE}
mixOmics::plotIndiv(getMixOmics(testObject, response = "temperature"))
```


